---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Introduction
This R markdown file is a literate programming document that anwsers questions 
in the Peer Assessment 1 of the Coursera course Reproducible Research. 

We follow the following conventions regarding citation:

* File names are put between "".
* Names of R objects are *italicized*

We set the global options so that every code chunk is visible.
```{r setoptions,echo=TRUE}
library(knitr)
opts_chunk$set(echo=TRUE)
```

## Loading and preprocessing the data
It is important that we make sure that this R markdown file and the data file "activity.csv" are in the same work directory.

We load the data into a data frame *initialData*. 
```{r}
initialData <- read.csv("activity.csv")
```

We check the structure and summary of the data frame *initialData*.
```{r}
str(initialData)
summary(initialData)
```

## What is mean total number of steps taken per day?

We use the dplyr package to compute the total steps of each day.
```{r,message=FALSE}
library(dplyr)
```


```{r}
stepsByDay <- initialData %>% group_by(date) %>% summarize(total_steps = sum(steps, na.rm =T))
stepsByDay
```

```{r,warning=FALSE}
library(ggplot2)

qplot(total_steps, data=stepsByDay, xlab="total steps in one day",binwidth = 1000)

```

The following is a histogram of the total number of steps taken each day.
```{r,message=FALSE}
steps <- as.numeric(stepsByDay$total_steps)
mean <- mean(steps)
median <- quantile(steps, probs=0.5)
```

The mean is `r round(mean, digits=0)`, and the median is `r format(round(median), scientific=F)`.


## What is the average daily activity pattern?





## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
